$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-03-15 qdpm_paper.tex
some minor changes. mostly fixed some typos.

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ and represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. ${T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $ {T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general, $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

@@

Because of the local constraints, all states $\ket{\psi}$ in the Hilbert space of a GQDM are invariant under local gauge transformations $G_v \ket{\psi} = \ket{\psi}$ the vertices, $G_v$ of the general form~\cite{Moessner2001}
\begin{equation}
G_v = e^{i \alpha_v Q_v}, \label{eq:Gv}
\end{equation}
where the allowed values of $\alpha_v$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

@@

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $\tilde{n}_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_0$ ($-c_0$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

@@

The two unique fixed-number constraints on the square lattice lead to QDM in the odd sector ($n_v = 1$) and the quantum full-packed loop model (QFPLM) in the even sector ($n_v = 2$). For configurations that satisfy these number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry~\cite{Moessner2001}.

@@

We have demonstrated that ground state at the RK point of the QDPM has an exact $Z_3$ gauge symmetry and a corresponding $9$-fold topological degeneracy on the torus. We emphasize that despite the exact $Z_3$ local gauge symmetry of the QDPM, there is not an exact mapping to a $Z_3$ gauge theory on the square lattice, as the QDPM has Ising degrees of freedom on the links, rather than the 3 dimensional local Hilbert space of a $Z_3$ gauge theory (see \appref{sec:Z3GT} for details of $Z_3$ gauge theory).  It remains to address whether this RK state is in the deconfined $Z_3$ topologically ordered phase of a $Z_3$ gauge theory, or a confined symmetry broken or critical state, which we discuss below.

@@

We explicitly demonstrate that the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$, which is shown in \figref{fig:u1_wind_qdpm}. The different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$, and we find that $\Phi_X$ and $\Phi_Y$ are not  strictly conserved (as they are in the QDM), but are only conserved modulo $3$. The histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors. For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ topological winding numbers $\Phi_X$ and $\Phi_Y$ as measured in for the ground state of the RK point of the QDPM for three different lattice sizes. The colors correspond to calculations initialized to different topological sectors $(\Phi_X,\Phi_Y)$: blue $(0,0)$, red $(1,1)$, and purple $(-1,-1)$. The lines are fits to a Gaussian for each system size. Note that $\Phi$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the $Z_3$ gauge theory description of the QDPM RK ground state, we study the $Z_3$ magnetic string correlation function as defined in \eqref{eq:Sgamma} \& \eqref{eq:SQDPM}:
\begin{equation}
C_M \left( \bm{r}_{pp'} \right) \equiv \left \langle \exp \left[ \frac{2 \pi}{3} i \left( \Phi_{\gamma_{pp'}} -\tilde{\Phi}_{\gamma_{pp'}} \right) \right] \right \rangle
\end{equation}
where $p$ and $p'$ are two plaquettes, $\bm{r}_{pp'}$ is the displacement between $p$ and $p'$,  and $\gamma_{pp'}$ is any open string on the dual lattice
connecting $p$ and $p'$. In \figref{fig:vison_3way} (a), we have plotted $|C_M|$ for a $L=32$ lattice. Note that while $C_M$ depends on the choice of the background dimerization which fixes the gauge, $|C_M|$ is independent of the background dimerization. We find only short range magnetic string correlations correlations that display no obvious symmetry breaking.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_z3_vis.pdf}
    \caption{ The magnitude of the $Z_3$ magnetic string correlation $|C_M|$ along along one axis of the dual lattice of a $L=64$ lattice. The line corresponds to an exponential fit with correlations length $\chi\approx 3$.}
    \label{fig:vison_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
%\begin{table}
%    \begin{tabular}{| l | c | c |}
%  	\hline
%     & $c$ & $\bar{c}$ \\
%	\hline
%    $i$ & $\monomermvA$ &  $\monomermvB$  \\
%    \hline
%  \end{tabular}
%  \caption{The pair of local dimer \& monomer configurations $c\leftrightarrow\bar{c}$ generate
%  minimal local monomer dynamics. Monomers always remain on their origonal sublattice. }
%  \label{tab:pentamerMoves}
%\end{table}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlations function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to $L=64$. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color.}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. Our numerical calculations explicitly show how the extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally, we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state. These results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liquid.

@@

A finite gap above the RK point suggests that a stable topologically ordered phase should exist surrounding the RK point. While the Monte Carlo techniques used in this manuscript are limited to the RK point, the phase diagram near the RK point can be explored in future work using quantum Monte Carlo methods, as the QDPM has no sign problem. Away from the RK point, the condensation of $Z_3$ magnetic vortices could drive a continuous phase transition out of the topologically order phase, analogous to vison condensation in $Z_2$ spin liquids~\cite{Jalabert1991,Ralko2007,Huh2011,Hao2014,Slagle2014}; if such a continuous transition exists in the QDPM it may provide access to a new universality of class of quantum transitions~\cite{XU2012}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$Z_3$ Gauge theory on the square lattice}
\label{sec:Z3GT}
Here we will review a string-net implementation of $Z_3$ gauge theory on the square lattice~\cite{Horn1979}. We define a three -dimensional Hilbert space on each link, corresponding to the presence or absence of an oriented string segment that carries $Z_3$ electric flux which are eigenstates of the Hermitian $Z_3$ electric flux operator $E_{vv'}$
\begin{align}
        & E_{vv'}  \eket{\vvemptylink} = 0
        ,
        \\
        & E_{vv'}  \eket{\rightarrowlink} = + \eket{\rightarrowlink}
        ,
        \\
        & E_{vv'} \eket{\leftarrowlink} = - \eket{\leftarrowlink}
        .
\end{align}

@@

We can define two oriented open string operators which commute with $H_{Z_3}$ everywhere except at the endpoints. The electric string operator $S_e$, which acts on an open, oriented string on the lattice $\gamma_{vv'}$ with endpoints $v,v'$, and creates a $+1$ ($-1$) charge at vertex $v$ ($v')$, is defined as
\begin{equation}
    S_e \left( \gamma_{vv'} \right) \equiv \prod_{l \in \gamma_{vv'}} Q_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ in $\gamma_{vv'}$, $v_l$ and $v_l'$ are the two vertices in $l$, ordered according to the orientation of $\gamma_{vv'}$ (see \figref{fig:example_elec_string}). The magnetic string operator $S_m$ is defined by an open, oriented string on the dual lattice and creates a magnetic vortex-antivortex pair on plaquettes at the endpoints of $\tilde{\gamma}_{pp'}$
\begin{equation}
    S_m \left( \tilde{\gamma}_{vv'} \right) \equiv \prod_{l \perp \tilde{\gamma}_{pp'}} P_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ that cross $\tilde{\gamma}$ and $v_l$ ($v_l'$) is the vertex to the left (right) of the string as $\tilde{\gamma}$ is traversed from $p$ to $p'$. 
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\linewidth]{example_elec_string.pdf}
    \caption{ ({\it left}) An example of a charge free $Z_3$ electric string configuration. The electric string operator $S_\gamma^e$ represented by the thick gray line, oriented from the white to black vertex. ({\it right}) An example of a $Z_3$ electric string configuration with a $+1$ ($-1$) electric charge at the red (blue) vertex; this string configuration results form the action of $S_\gamma^e$ on the left panel.  }
    \label{fig:example_elec_string}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}
%    \centering
%    \includegraphics[width=1.0\columnwidth]{spatial_cors_perp.pdf}
%    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
%    sub-lattice for a $L=64$ lattice.}
%    \label{fig:spatial_dmr_cor}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

Note that  $S_e$ and $S_m$ do not commute if their corresponding strings intersect--this underlies the non-trivial statics of electric charges and magnetic vortices. Consider the magnetic string operator $S_m(\tilde{\Gamma})$ defined by an oriented \emph{closed} loop $\tilde{\Gamma}$ on the dual lattice. This operator corresponds to the creation of a vortex-antivortex pair, winding of the vortex around $\tilde{\Gamma}$, and annihilation of the vortex pair. A state with definite electric charge inside $\tilde{\Gamma}$ will be an eigenstate of $S_m(\tilde{\Gamma})$, with eigenvalue $e^{ i 2 \pi/3 q}$, where $q=0,\pm1$ is the net $Z_3$ electric charge (modulo 3) inside $\tilde{\Gamma}$. Thus for $\tilde{\Gamma}$ winding around an isolated electric charge, we see that electric charges and magnetic vortices are anyons with relative braiding statics with a statistical phase of $e^{\pm i 2\pi/3}$.

>>>>

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu  {T}_{\nu} + \sum_\nu v_\nu  {V}_{\nu} \label{eq:HGQDM}
\end{equation}
where ${T}_\nu$ are off-diagonal kinetic energy operators, ${V}_\nu$ are diagonal potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ represents a set of  neighboring links. We will limit our discussion to $t_\nu >0$. ${T}_\nu$ is constructed to generate resonances between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
 {T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $ {T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
 {V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
Therefore we may define a Hamiltonian for a GQDM using \Eqref{eq:HGQDM} and defining a set of dimer resonances $\{(c_\nu,\bar{c}_\nu)\}$ and their corresponding strengths $\{(t_\nu,v_\nu)\}$:
\begin{equation}
\HGQDM \biggl( \bigl\{\left(c_\nu,\bar{c}_\nu\right)\},\{\left(t_\nu,v_\nu\right)\bigr\}\biggr).
\end{equation}
 In general, $[T_\nu, T_{\nu'}] \neq 0 $ and $[T_\nu,V_{\nu'}]\neq0$ when $\nu$ and $\nu'$ share links, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

@@

Because of the local constraints, all states $\ket{\psi}$ in the Hilbert space of a GQDM are invariant under local gauge transformations $G_v \ket{\psi} = \ket{\psi}$, $G_v$ of the general form~\cite{Moessner2001}
\begin{equation}
G_v = e^{i \alpha_v Q_v}, \label{eq:Gv}
\end{equation}
where the allowed values of $\alpha_v$ depend on the constraint.  
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

@@

The gauge symmetry allows for the construction of a gauge invariant magnetic string operator $S_\gamma$ along an oriented path $\gamma$ path on the dual lattice. We first orient all the links of the square lattice such that they point from the $A$ sublattice to the $B$ sublattice (see \figref{fig:orientedSL}) and define an oriented dimer flux $\Phi_\gamma$ through the path:
\begin{equation}
\Phi_\gamma \equiv \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) n_\ell \label{eq:phi}
\end{equation}
where $\mathrm{sgn}(\ell,\gamma)$ is positive (negative) if $\ell$ is oriented to the right (left) with respect to the orientation of $\gamma$. Additionally, we define background flux $\tilde{\Phi}_\ell$
\begin{equation}
\tilde{\Phi}_\gamma \equiv  \sum_{\ell \in \gamma} \rm{sgn}\left(\ell,\gamma\right) \tilde{n}_\ell \label{eq:phiBackground}
\end{equation}
where $\tilde{n}_\ell$ is the number operator for a static background dimer configuration which is chosen to fix the gauge. With this gauge fixing, the total charge enclosed in a closed loop $\Gamma$ is $Q_\Gamma = \Phi_\Gamma- \tilde{\Phi}_\Gamma$. An alternative physical picture is that there are static charges of charge $+c_0$ ($-c_0$) sitting at the vertices of the $A$($B$) sub-lattice, which are the source of the non-vanishing $\Phi_\Gamma$ flux through a closed loop.

@@

The two unique fixed-number constraints on the square lattice lead to the QDM in the odd sector ($n_v = 1$) and the quantum full-packed loop model (QFPLM) in the even sector ($n_v = 2$). For configurations that satisfy these number constraints, $Q_v$ strictly vanishes, and thus all states in the Hilbert space are invariant under $G_v$ for $0 \leq \alpha_v \leq 2 \pi$ in \Eqref{eq:Gv}; this represents a local $U(1)$ gauge symmetry~\cite{Moessner2001}.

@@

We have demonstrated that the ground state at the RK point of the QDPM has an exact $Z_3$ gauge symmetry and a corresponding $9$-fold topological degeneracy on the torus. We emphasize that despite the exact $Z_3$ local gauge symmetry of the QDPM, there is not an exact mapping to a $Z_3$ gauge theory on the square lattice, as the QDPM has Ising degrees of freedom on the links, rather than the 3 dimensional local Hilbert space of a $Z_3$ gauge theory (see \appref{sec:Z3GT} for details of $Z_3$ gauge theory).  It remains to address whether this RK state is in the deconfined $Z_3$ topologically ordered phase of a $Z_3$ gauge theory, or a confined symmetry broken or critical state, which we discuss below.

@@

We explicitly demonstrate that the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$, which is shown in \figref{fig:u1_wind_qdpm}. The different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$, and we find that $\Phi_X$ and $\Phi_Y$ are not  strictly conserved (as they are in the QDM), but are only conserved modulo $3$. The histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors. For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ topological winding numbers $\Phi_X$ and $\Phi_Y$ as measured in the ground state of the RK point of the QDPM for three different lattice sizes. The colors correspond to calculations initialized to different topological sectors $(\Phi_X,\Phi_Y)$: blue $(0,0)$, red $(1,1)$, and purple $(-1,-1)$. The lines are fits to a Gaussian for each system size. Note that $\Phi$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

To investigate the $Z_3$ gauge theory description of the QDPM RK ground state, we study the $Z_3$ magnetic string correlation function as defined in \eqref{eq:Sgamma} \& \eqref{eq:SQDPM}:
\begin{equation}
C_M \left( \bm{r}_{pp'} \right) \equiv \left \langle \exp \left[ \frac{2 \pi}{3} i \left( \Phi_{\gamma_{pp'}} -\tilde{\Phi}_{\gamma_{pp'}} \right) \right] \right \rangle
\end{equation}
where $p$ and $p'$ are two plaquettes, $\bm{r}_{pp'}$ is the displacement between $p$ and $p'$,  and $\gamma_{pp'}$ is any open string on the dual lattice
connecting $p$ and $p'$. In \figref{fig:vison_3way} (a), we have plotted $|C_M|$ for a $L=32$ lattice. Note that while $C_M$ depends on the choice of the background dimerization which fixes the gauge, $|C_M|$ is independent of the background dimerization. We find only short range magnetic string correlations that display no obvious symmetry breaking.

@@

%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includegraphics[width=1.0\columnwidth]{spatial_cors_z3_vis.pdf}
    \caption{ The magnitude of the $Z_3$ magnetic string correlation $|C_M|$ along one axis of the dual lattice of a $L=64$ lattice. The line corresponds to an exponential fit with correlations length $\chi\approx 3$.}
    \label{fig:vison_cor}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
%\begin{table}
%    \begin{tabular}{| l | c | c |}
%  	\hline
%     & $c$ & $\bar{c}$ \\
%	\hline
%    $i$ & $\monomermvA$ &  $\monomermvB$  \\
%    \hline
%  \end{tabular}
%  \caption{The pair of local dimer \& monomer configurations $c\leftrightarrow\bar{c}$ generate
%  minimal local monomer dynamics. Monomers always remain on their origonal sublattice. }
%  \label{tab:pentamerMoves}
%\end{table}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[]
        \centering
        \includegraphics[width=1.0\columnwidth]{monomer_cor_log.pdf}
        \caption{ The monomer-monomer correlation function $\langle m_{\bm{0}} m_{\bm{r}} \rangle$ along a coordinate axis in the QDPM RK ground state doped with a single pair of monomers on lattices up to $L=64$. The horizontal lines correspond to $L^{-4}$ for the system size corresponding to the color.}
        \label{fig:monomer_correlations}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid without spontaneously broken symmetry. Our numerical calculations explicitly show how the extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally, we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state. These results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liquid.

@@

A finite gap above the RK point suggests that a stable topologically ordered phase should exist surrounding the RK point. While the Monte Carlo techniques used in this manuscript are limited to the RK point, the phase diagram near the RK point can be explored in future work using quantum Monte Carlo methods, as the QDPM has no sign problem. Away from the RK point, the condensation of $Z_3$ magnetic vortices could drive a continuous phase transition out of the topologically order phase, analogous to vison condensation in $Z_2$ spin liquids~\cite{Jalabert1991,Ralko2007,Huh2011,Hao2014,Slagle2014}; if such a continuous transition exists in the QDPM it may provide access to a new universality class of quantum transitions~\cite{XU2012}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$Z_3$ Gauge theory on the square lattice}
\label{sec:Z3GT}
Here we will review a string-net implementation of $Z_3$ gauge theory on the square
lattice~\cite{Horn1979}. We define a three -dimensional Hilbert space on each link, corresponding to the presence or absence of an oriented string segment that carries $Z_3$ electric flux, which are eigenstates of the Hermitian $Z_3$ electric flux operator $E_{vv'}$
\begin{align}
        & E_{vv'}  \eket{\vvemptylink} = 0
        ,
        \\
        & E_{vv'}  \eket{\rightarrowlink} = + \eket{\rightarrowlink}
        ,
        \\
        & E_{vv'} \eket{\leftarrowlink} = - \eket{\leftarrowlink}
        .
\end{align}

@@

We can define two oriented open string operators which commute with $H_{Z_3}$ everywhere except at the endpoints. The electric string operator $S_e$, which acts on an open, oriented string on the lattice $\gamma_{vv'}$ with endpoints $v,v'$, and creates a $+1$ ($-1$) charge at vertex $v$ ($v')$, is defined as
\begin{equation}
    S_e \left( \gamma_{vv'} \right) \equiv \prod_{l \in \gamma_{vv'}} Q_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ in $\gamma_{vv'}$, $v_l$ and $v_l'$ are the two vertices in $l$, ordered according to the orientation of $\gamma_{vv'}$ (see \figref{fig:example_elec_string}). The magnetic string operator $S_m$ is defined by an open, oriented string on the dual lattice and creates a magnetic vortex-antivortex pair on plaquettes at the endpoints of $\tilde{\gamma}_{pp'}$
\begin{equation}
    S_m \left( \tilde{\gamma}_{vv'} \right) \equiv \prod_{l \perp \tilde{\gamma}_{pp'}} P_{v_l v_l'}^\dagger
\end{equation}
where the product is over all links $l$ that cross $\tilde{\gamma}$ and $v_l$ ($v_l'$) is the vertex to the left (right) of the string as $\tilde{\gamma}$ is traversed from $p$ to $p'$. 
%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\linewidth]{example_elec_string.pdf}
    \caption{ ({\it left}) An example of a charge free $Z_3$ electric string configuration. The electric string operator $S_\gamma^e$ is represented by the thick gray line, oriented from the white to black vertex. ({\it right}) An example of a $Z_3$ electric string configuration with a $+1$ ($-1$) electric charge at the red (blue) vertex; this string configuration results form the action of $S_\gamma^e$ on the left panel.  }
    \label{fig:example_elec_string}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}
%    \centering
%    \includegraphics[width=1.0\columnwidth]{spatial_cors_perp.pdf}
%    \caption{Correlation function between perpendicular dimers on the same (blue) and different (red) 
%    sub-lattice for a $L=64$ lattice.}
%    \label{fig:spatial_dmr_cor}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

Note that  $S_e$ and $S_m$ do not commute if their corresponding strings intersect--this underlies the non-trivial statistics of electric charges and magnetic vortices. Consider the magnetic string operator $S_m(\tilde{\Gamma})$ defined by an oriented \emph{closed} loop $\tilde{\Gamma}$ on the dual lattice. This operator corresponds to the creation of a vortex-antivortex pair, winding of the vortex around $\tilde{\Gamma}$, and annihilation of the vortex pair. A state with definite electric charge inside $\tilde{\Gamma}$ will be an eigenstate of $S_m(\tilde{\Gamma})$, with eigenvalue $e^{ i 2 \pi/3 q}$, where $q=0,\pm1$ is the net $Z_3$ electric charge (modulo 3) inside $\tilde{\Gamma}$.  Thus for $\tilde{\Gamma}$ winding around an isolated electric charge, we see that electric charges and magnetic vortices are anyons with relative braiding statistics with a statistical phase of $e^{\pm i 2\pi/3}$.
